---
title: 'Human Activity Recognition : A Qualitative Assessment of Weight Lifting Exercises'
author: "Abhinav Mishra"
date: "21 February 2015"
output: html_document
bibliography: bibliography.bib
---
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, we use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants [@velloso2013].

They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. Given data from accelerometers, the goal is to predict the class of action which is one of the following.

- exactly according to the specification (A)
- throwing elbows to the front (B)
- lifting the dumbbell only halfway (C)
- lowering the dumbbell only halfway (D)
- throwing the hips to the front (E).

More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

# Data
The training data for this project are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv  
The test data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv  

## Loading train data and test data
```{r loading_data, cache=TRUE}
library(RCurl)
train_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
train_data <- read.csv(text=getURL(train_url), na.strings=c("", "NA"))
test_data <- read.csv(text=getURL(test_url), na.strings=c("", "NA"))
```

## Preprocessing for consumption
```{r preprocess, message=FALSE}
# 1. Remove first column of the data (which is just index and of no relevance).
train_data$X <- NULL

# 2. Remove the user and time information (which do not have any effect on whether barbell lifts are performed correctly or not).
ineffective_cols <- c("user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp")
for (col in ineffective_cols) {
    train_data[, col] <- NULL
}

# 3. Remove the columns which have many missing values and imputing is not possible.
NAs <- apply(train_data,2,function(x) {sum(is.na(x))})
train_data <- train_data[,which(NAs == 0)]

# 4. Remove features that don't have any (or very small)  variance predictors
library(caret)
novar <- nearZeroVar(train_data)
train_data <- train_data[-novar]
test_data <- test_data[-novar]
```

The final set of predictors/ features used for classification are:-
```{r readydata}
names(train_data)
```

# Modellling
We build a random forest classifier to predict the action class. Let's measure the accuracy of the model
Perform a 10-fold cross validation with 80:20 split, on each fold, use 80% of the data for training the random forest and use remaining 20% for testing.
```{r cache=TRUE}
library(randomForest)
set.seed(1)
obs <- c()
preds <- c()
for(i in 1:10) {
    intrain = sample(1:dim(train_data)[1], size=dim(train_data)[1] * 0.8, replace=F)
    train_cross = train_data[intrain,]
    test_cross = train_data[-intrain,]
    rf <- randomForest(classe ~ ., data=train_cross)
    obs <- c(obs, test_cross$classe)
    preds <- c(preds, predict(rf, test_cross))
}
```

The confusion matrix:-
```{r}
conf_mat <- confusionMatrix(table(preds, obs))
conf_mat$table
```

The accuracy is `r conf_mat$overall[[1]] * 100`% and only few instances are wrongly classified. So the model seems good. 
Train the random forest with whole dataset so that the classifier can be used to predict the class of an action, given the set of activity measurements.
```{r cache=TRUE}
model <- randomForest(classe ~ ., data=train_data)
```

# References
